Mathematical Approaches for Simulating Epidemic Progression: Addressing Limitations of the Linear Chain Trick in ODE Models

### 2024-04-23 thesis/draft.tex
some changes for thesis

In the field of epidemiology, ordinary differential equations (ODEs) serve as a crucial tool for characterizing disease progression and making predictions. During the COVID-19 pandemic, they were extensively utilized to guide policy-making, especially in disease control and public health interventions \cite{thompson2020epidemiological}. A commonly used model is the SIR compartmental model \cite{Anderson1991}\cite{kermack1927contribution}, which divides the population into three states: Susceptible $(S)$, Infectious $(I)$, and Recovered $(R)$. In the model, a susceptible individual, upon contact with the infected population at a rate $\beta$ $[\frac{1}{\mathit{ time}}]$, transitions immediately to the infectious state. An infected individual then moves to the recovered state at a recovery rate of $\gamma$ $[\frac{1}{\mathit{time}}]$. A major limitation of this model is its implicit assumption that the duration of the infectious stage follows an exponential distribution. This can be observed in a simplified equation that models the dynamics of an infected cohort. The proportion of a cohort infected at the same time begins at 1, and follows:
\begin{align}
    \frac{dI}{dt} = -\gamma I
\end{align}
Solving this equation explicitly yields
\begin{align}
    I(t) = e^{-\gamma t}.
	 \label{explicitCohort}
\end{align}
\Eqnref{explicitCohort} implies a stochastic state transition model where the time spend in the infectious $I$ is exponentially distributed with a mean stage duration of $\frac{1}{\gamma}$. \jd{Not really. Where did the stochastic part come from? Also, these things are implied by the assumptions, not the calculation.}
\Eqnref{explicitCohort} implies that an individual has the highest probability of recovering from the disease immediately after infection, which is generally not realistic. In reality, a delay between the successful infection of an individual and the recovery from the disease is inevitable. The work conducted by Byrne AW et al.~\cite{byrne2020inferred} presents the simulation of parameter distribution inferred for the duration of the infectious period, and the shape of this distribution closely resembles a gamma distribution. The study by Greenhalgh and Day (2017) \cite{greenhalgh2017time}, utilizing measles outbreak data, also demonstrates that the recovery rate from an infectious disease is not constant but can vary based on a wide range of factors. \jd{We're not really interested in factors, though, only in time-dependence. Is this citation good for that?}

Even when the mean is constant, the shape of the stage duration distributions (i.e., the timing of stage transitions) can significantly impact the dynamics and the outcomes of predictive models in applied settings \cite{krylova2013effects}\cite{keeling2002understanding}\cite{wearing2005appropriate}\cite{nguyen2008noise}. Consequently, it becomes crucial to consider the type of distribution characterizing an individual in a specific state when constructing these models.\jd{Try to be more consistent (and simpler) about what you call these distributions. Dwell-time, characterizing individual in state, stage-duration, are all saying the same thing.} A standard approach for developing models that incorporate different stage duration distributions is through the use of integro-differential equations (IDEs) or integral equations (IEs) \cite{hurtado2019generalizations}\cite{kermack1927contribution}\cite{hethcote1980integral}. IDEs facilitate the extension of the SIR model to include arbitrary distributions for the duration of infectiousness, as discussed in various sources \cite{feng2000endemic}\cite{hethcote1980integral}\cite{ma2006generality}. This approach allows for the capture of the `memory effect', where the current rate of change in each compartment of the model (Susceptible, Infectious, Recovered) is influenced by the entire history of the epidemic, rather than relying on a constant rate assumption.

@@

In this study, we focused on one particular direction: generalizing the Erlang distribution of stage durations currently present in the model. The goals were to have a more effective way for parameter estimation and to eliminate the constraints imposed on the shape parameter ($\kappa$). To address these limitations, we proposed the Geometric Chain Trick (GCT). Instead of having a variable number of substages $(n)$ with a constant transition rate between them $(\gamma_i = n\gamma)$, we fix $n$ to a specific integer value and use a series of rates that follow a geometric pattern. This series is represented by the formula $\gamma_i = ar^{i-1}$, governed by parameters $a$ and $r$. This approach enables a uniform model structure with the parameter of interest transitioning from discrete values $(n$ and $n\gamma)$ to continuous values $(a$ and $r)$, thereby enhancing the model's adaptability (details in \sref{Methods}).

To illustrate this point, consider that the Gamma distribution can be parameterized in terms of its mean $(D)$ and shape parameter (quadratic coefficient of variation, $\kappa$), which determine the overall shape and position of the distribution. As mentioned above, in subdividing a single stage into n substages, each exponentially distributed substage must have a rate of $n\gamma$ to maintain the overall rate of that stage. While the mean of the Erlang distribution in the ODE system can be chosen freely, $\kappa$ is constrained to discrete values, shown as follows:
\begin{align*}
    Mean: \quad &\sum_{i=1}^{n_f} \frac{1}{n\gamma} = \frac{1}{\gamma}\\
    Variance: \quad &\sum_{i=1}^{n_f} \frac{1}{n^2\gamma^2} = \frac{1}{n\gamma^2} \\
    \kappa: \quad &\frac{\sigma^2}{M^2} = \frac{1}{n} \quad n \in \mathbb{N}
\end{align*}
Our approach, using a geometric series of rates, addresses this issue by allowing $\kappa$ to take any value within the range $\left[ \frac{1}{n}, 1 \right)$. A detailed explanation and the derivation process are provided in Section $4.1$. Another advantage of this method is that it enables a fixed model structure where only the parameters change, eliminating the need to determine optimal substage numbers through the construction of multiple models, a process currently required in the ODE model that employs the LCT.

@@

\section{Methods}
\label{Methods}
Several assumptions have been made for the unforced SIR model (Susceptible, Infectious, and Recovered) that our study is based on: 
\begin{enumerate}
    \item \textbf{Closed Population}: A constant population size with equal birth and death rates and no disease-related mortalities.
    \item \textbf{Homogeneous mixing}: Individuals have an equal chance of coming into contact with any other individual in the population. 
    \item \textbf{Ignoring Latency}: Individuals move immediately from a susceptible state to an infected state, ignoring any incubation period associated with the disease.
    \item \textbf{Lifelong immunity}: Once an individual recovers, they do not become susceptible or infectious again.
    \item \textbf{Fixed rate of transmission and recovery}: The rate of disease spread and the rate at which infected individuals recover and become immune are constant over time.
\end{enumerate}

@@

\section{Results}
\subsection{Independent and Flexible Modification of Mean and Kappa in Stage Duration Distribution Using the GCT Model}
\jd{What is new in this subsection? Can it be merged into Methods?}
As discussed in previous sections, the duration distribution of the stage can be parameterized by its mean (D) and shape parameter $(\kappa^2 = CV^2)$. In the GCT model, the calculation of mean and variance is achieved by summing the respective properties of each substage, shown as follows:
\begin{align}
    Mean: \quad  &\sum_{i=1}^{n_f} \frac{1}{\gamma_i} = \frac{\frac{1}{a} (\frac{1}{r^n} - 1)}{\frac{1}{r} -1} \\
    Variance: \quad &\sum_{i=1}^{n_f} \left(\frac{1}{\gamma_i} \right)^2 = \frac{\frac{1}{a^2} (\frac{1}{r^{2n}} - 1)}{\frac{1}{r^2} -1}
\end{align}
The shape parameter $\kappa$, determined by $\frac{\sigma^2}{D}$, can thus be obtained:
\begin{align}
    \kappa = \frac{(1+r^{n}) (1-r)}{(1+r) (1-r^{n})}
\end{align}
This simplification revealed that the value of $\kappa$ is solely determined by the parameter $r$, thus enabling the alignment of kappa without affecting the mean, which can be adjusted through alterations to parameter $a$. Similarly, the mean can be scaled to a desired value without impacting $\kappa$. This grants the model independence in modifying properties.

@@

Using formulas (18) and (20), we developed code to reverse-engineer the process: given specific values of D and kappa, we can identify the corresponding $a$ and $r$ that endow the model with such characteristics. These $a$ and $r$ values were subsequently integrated into the GCT model. Our initial goal was to determine if the stage duration distribution could match between the two models, assuming identical $D$ and $\kappa$ values. The number of substages in the LCT model was set to 2, 4, 6, and 8, while the substage number for the GCT model was fixed at 12. We selected a mean of 10 days. Applying the mean and kappa values $\left( \frac{1}{2}, \frac{1}{4}, \frac{1}{6}, \frac{1}{8} \right)$ to the system, we plotted the stage duration distributions generated from the LCT cohort model alongside those from the GCT with the corresponding a and r values. As depicted in Figure \ref{distributionCompare}, the stage duration distributions exhibit slightly different shapes, despite having identical mean and shape parameters. To further examine the match of properties, D and kappa were numerically calculated from the output distributions, revealing identical values across both models. This indicates that while the distributions' shapes may not completely match, they can possess the same properties.

@@

\subsection{Parameter Estimation}
In this section, we discuss the parameter estimation results obtained through model fitting. The incidence data were generated using the GCT model with 5 substages, setting the $D$ to 10 and $\kappa$ to $\frac{2}{9}$. The model was fitted using the GCT framework with 5 to 10 substages. As depicted in Figure \ref{parEst}, the precise values for each parameter are marked by a red line, with post-fitting parameter values displayed alongside their 95\% confidence intervals. It can be observed that the parameter estimation is not perfect, with kappa and beta consistently overestimated, and D underestimated. However, these discrepancies in parameter estimation counterbalance each other, resulting in a fitting outcome characterized by a low mean squared error. It is possible for the parameter to lack confidence intervals, which could be attributed to structural inadequacies. As previously mentioned, the minimum possible value for kappa in the model with $n$ substages is $\frac{1}{n}$. The model encounters limitations when aligning with a kappa value of less than $\frac{1}{n}$. Overall, it is important to acknowledge the imprecision of parameter estimation, which will be a focus of our subsequent efforts.

\begin{figure}[h!]
    \centering
    \includegraphics[width= \textwidth]{figures/parEst.Rout.pdf}
    \caption{\textbf{GCT parameter estimation for various structures.} \\ Input data for fittings were derived from the GCT transmission model featuring 5 subtages, with $\beta=0.2$, $\mu=0.01$, $N=10000$, $I_0=10$. Fittings were conducted for 5 to 10 substages within the GCT framework. The targeted parameter values are marked by red lines, with the 95\% confidence intervals depicted by black lines.}
    \label{parEst}
\end{figure}

>>>>

In the field of epidemiology, ordinary differential equations (ODEs) serve as a crucial tool for characterizing disease progression and making predictions. During the COVID-19 pandemic, they were extensively utilized to guide policy-making, especially in disease control and public health interventions \cite{thompson2020epidemiological}. A commonly used model is the SIR compartmental model \cite{Anderson1991}\cite{kermack1927contribution}, which divides the population into three states: Susceptible $(S)$, Infectious $(I)$, and Recovered $(R)$. In the model, a susceptible individual, upon contact with the infected population at a rate $\beta$ $[\frac{1}{\mathit{ time}}]$, transitions immediately to the infectious state. An infected individual then moves to the recovered state at a recovery rate of $\gamma$ $[\frac{1}{\mathit{time}}]$. A major limitation of this model is its implicit assumption that the duration of the infectious stage follows an exponential distribution. This can be observed in a simplified equation that models the dynamics of an infected cohort. The proportion of a cohort infected at the same time begins at 1, and follows:
\begin{align}
    \frac{dI}{dt} = -\gamma I
\end{align}
Solving this equation explicitly yields
\begin{align}
    I(t) = e^{-\gamma t}.
	 \label{explicitCohort}
\end{align}
In other words, the time spent in the infectious stage $I$ is exponentially distributed with a mean stage duration of $\frac{1}{\gamma}$.
\Eqnref{explicitCohort} implies that an individual has the highest probability of recovering from the disease immediately after infection, which is generally not realistic. In reality, a delay between the successful infection of an individual and the recovery from the disease is inevitable. The work conducted by Byrne AW et al.~\cite{byrne2020inferred} presents the simulation of parameter distribution inferred for the duration of the infectious period, and the shape of this distribution closely resembles a gamma distribution.

Even when the mean is constant, the shape of the stage duration distributions (i.e., the timing of stage transitions) can significantly impact the dynamics and the outcomes of predictive models in applied settings \cite{krylova2013effects}\cite{keeling2002understanding}\cite{wearing2005appropriate}\cite{nguyen2008noise}. Consequently, it becomes crucial to consider the type of distribution characterizing an individual in a specific state when constructing these models. A standard approach for developing models that incorporate different stage duration distributions is through the use of integro-differential equations (IDEs) or integral equations (IEs) \cite{hurtado2019generalizations}\cite{kermack1927contribution}\cite{hethcote1980integral}. IDEs facilitate the extension of the SIR model to include arbitrary distributions for the duration of infectiousness, as discussed in various sources \cite{feng2000endemic}\cite{hethcote1980integral}\cite{ma2006generality}. This approach allows for the capture of the `memory effect', where the current rate of change in each compartment of the model (Susceptible, Infectious, Recovered) is influenced by the entire history of the epidemic, rather than relying on a constant rate assumption.

@@

In this study, we focused on one particular direction: generalizing the Erlang distribution of stage durations currently present in the model. The goals were to have a more effective way for parameter estimation and to eliminate the constraints imposed on the shape parameter ($\kappa$). To address these limitations, we proposed the Geometric Chain Trick (GCT). Instead of having a variable number of substages $(n)$ with a constant transition rate between them $(\gamma_i = n\gamma)$, we fix $n$ to a specific integer value and use a series of rates that follow a geometric pattern. This series is represented by the formula $\gamma_i = ar^{i-1}$, governed by parameters $a$ and $r$. This approach enables a uniform model structure with the parameter of interest transitioning from discrete values $(n$ and $n\gamma)$ to continuous values $(a$ and $r)$, thereby enhancing the model's adaptability (details in \sref{Models}).

To illustrate this point, consider that the Gamma distribution can be parameterized in terms of its mean $(D)$ and shape parameter (quadratic coefficient of variation, $\kappa$), which determine the overall shape and position of the distribution. As mentioned above, in subdividing a single stage into n substages, each exponentially distributed substage must have a rate of $n\gamma$ to maintain the overall rate of that stage. While the mean of the Erlang distribution in the ODE system can be chosen freely, $\kappa$ is constrained to discrete values, shown as follows:
\begin{align*}
    Mean: \quad &\sum_{i=1}^{n_f} \frac{1}{n\gamma} = \frac{1}{\gamma}\\
    Variance: \quad &\sum_{i=1}^{n_f} \frac{1}{n^2\gamma^2} = \frac{1}{n\gamma^2} \\
    \kappa: \quad &\frac{\sigma^2}{M^2} = \frac{1}{n} \quad n \in \mathbb{N}
\end{align*}
Our approach, using a geometric series of rates, addresses this issue by allowing $\kappa$ to take any value within the range $\left[ \frac{1}{n}, 1 \right)$. A detailed explanation and the derivation process are provided in \sref{parm result}. Another advantage of this method is that it enables a fixed model structure where only the parameters change, eliminating the need to determine optimal substage numbers through the construction of multiple models, a process currently required in the ODE model that employs the LCT.

@@

\section{Models}
\label{Models}
Several assumptions have been made for the unforced SIR model (Susceptible, Infectious, and Recovered) that our study is based on: 
\begin{enumerate}
    \item \textbf{Closed Population}: A constant population size with equal birth and death rates and no disease-related mortalities.
    \item \textbf{Homogeneous mixing}: Individuals have an equal chance of coming into contact with any other individual in the population. 
    \item \textbf{Ignoring Latency}: Individuals move immediately from a susceptible state to an infected state, ignoring any incubation period associated with the disease.
    \item \textbf{Lifelong immunity}: Once an individual recovers, they do not become susceptible or infectious again.
    \item \textbf{Fixed rate of transmission and recovery}: The rate of disease spread and the rate at which infected individuals recover and become immune are constant over time.
\end{enumerate}

@@

\section{Results}
\subsection{Independent and Flexible Modification of Mean and Kappa in Stage Duration Distribution Using the GCT Model}
\label{parm result}
As discussed in previous sections, the duration distribution of the stage can be parameterized by its mean (D) and shape parameter $(\kappa^2 = CV^2)$. In the GCT model, the calculation of mean and variance is achieved by summing the respective properties of each substage, shown as follows:
\begin{align}
    Mean: \quad  &\sum_{i=1}^{n_f} \frac{1}{\gamma_i} = \frac{\frac{1}{a} (\frac{1}{r^n} - 1)}{\frac{1}{r} -1} \\
    Variance: \quad &\sum_{i=1}^{n_f} \left(\frac{1}{\gamma_i} \right)^2 = \frac{\frac{1}{a^2} (\frac{1}{r^{2n}} - 1)}{\frac{1}{r^2} -1}
\end{align}
The shape parameter $\kappa$, determined by $\frac{\sigma^2}{D}$, can thus be obtained:
\begin{align}
    \kappa = \frac{(1+r^{n}) (1-r)}{(1+r) (1-r^{n})}
\end{align}
This simplification revealed that the value of $\kappa$ is solely determined by the parameter $r$, thus enabling the alignment of kappa without affecting the mean, which can be adjusted through alterations to parameter $a$. Similarly, the mean can be scaled to a desired value without impacting $\kappa$. This grants the model independence in modifying properties.

@@

Using formulas (18) and (20), we developed code to reverse-engineer the process: given specific values of D and kappa, we can identify the corresponding $a$ and $r$ that endow the model with such characteristics. These $a$ and $r$ values were subsequently integrated into the GCT model. Our initial goal was to determine if the stage duration distribution could match between the two models, assuming identical $D$ and $\kappa$ values. The number of substages in the LCT model was set to 2, 4, 6, and 8, while the substage number for the GCT model was fixed at 12. We selected a mean of 10 days. Applying the mean and kappa values $\left[ \frac{1}{2}, \frac{1}{4}, \frac{1}{6}, \frac{1}{8} \right]$ to the system, we plotted the stage duration distributions generated from the LCT cohort model alongside those from the GCT with the corresponding a and r values. As depicted in Figure \ref{distributionCompare}, the stage duration distributions exhibit slightly different shapes, despite having identical mean and shape parameters. To further examine the match of properties, D and kappa were numerically calculated from the output distributions, revealing identical values across both models. This indicates that while the distributions' shapes may not completely match, they can possess the same properties.

@@

\subsection{Parameter Estimation}
In this section, we discuss the parameter estimation results obtained through model fitting. The true incidence data were generated using the GCT model with 5 substages, setting the $D$ to 10 and $\kappa$ to $\frac{2}{9}$. Random noises were added to the true incidence data using negative binomial to give us 4 different input data. The model was fitted using the GCT framework with 5 to 10 substages. As depicted in Figure \ref{parEst}, the precise values for each parameter are marked by a red line, with post-fitting parameter values displayed alongside their 95\% confidence intervals. It can be observed that the parameter estimation is not perfect, with kappa and beta consistently overestimated, and D underestimated. However, these discrepancies in parameter estimation counterbalance each other, resulting in a fitting outcome characterized by a low mean squared error. It is possible for the parameter to lack confidence intervals, which could be attributed to structural inadequacies. As previously mentioned, the minimum possible value for kappa in the model with $n$ substages is $\frac{1}{n}$. The model encounters limitations when aligning with a kappa value of less than $\frac{1}{n}$. Overall, it is important to acknowledge the imprecision of parameter estimation, which will be a focus of our subsequent efforts.

\begin{figure}[h!]
    \centering
    \includegraphics[width= 1.1\textwidth]{figures/parEst.Rout.pdf}
    \caption{\textbf{GCT parameter estimation for various structures.} \\ Input data for fittings were derived from the GCT transmission model featuring 5 subtages, with $\beta=0.2$, $\mu=0.01$, $N=10000$, $I_0=10$. Fittings were conducted for 5 to 10 substages within the GCT framework. The targeted parameter values are marked by red lines, with the 95\% confidence intervals depicted by black lines.}
    \label{parEst}
\end{figure}
